Multitask matrix completion for learning protein interactions across diseases

### 2016-09-01 paper.tex
citations and results

\abstract{\textbf{Motivation:}
Disease causing pathogens such as viruses, introduce their proteins into the host cells 
where they interact with the host's proteins enabling the virus to replicate inside the host.
These interactions between pathogen and host proteins are key to understanding infectious diseases.
Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. \\
\textbf{Results:}
Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
We obtain between 7 to 39 percentage points improvement in predictive performance over prior state-of-the-art models. 
We show how our model's parameters can be interpreted to reveal both general and specific
interaction-relevant characteristics of the viruses. \\
\textbf{Availability:}
Our code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl\_mtl.tgz}\\
\textbf{Contact:}
\href{mkshirsa@cs.cmu.edu}{mkshirsa@cs.cmu.edu}
}

@@

\section{Dataset}
\label{sec:datasets}
We use three human-virus PPI datasets from the PHISTO \citep{phisto} database (version from 2014), the characteristics of which are summarized in Table \ref{datasets}. 
The \textit{Influenza A} task includes various strains of flu such as: H1N1 (Puerto Rico, Texas strains etc), H3N2 (England strain etc), H5N1, H7N3.
Similarly, the \textit{Hepatitis} task includes strains such as of the virus. 
All three are single-strand RNA viruses, with \textit{Hepatitis} being a positive-strand ssRNA whereas \textit{Influenza} and \textit{Ebola} are negative-strand viruses. The density of the known interactions is quite small when considering the entire proteome (i.e all known proteins) of the host and pathogen species (row-4 in Table \ref{datasets}). \\

\subsection{Addressing the homologs in PHISTO}
\label{sec:phisto}
The interactions data reported in PHISTO database is aggregated from several host-pathogen interactions data sources such as Virus-Mint, PHI-base etc. We found that many of the interactions (within a task) involve homologous PPI - virus protein homologs from two strains interacting with either the same human protein or with homologous human proteins. Such PPI usually come from different publications/ studies, but they introduce a lot of redundancy in the dataset. Removing such PPI is challenging as there are several criteria to consider, the most important being: \textit{what threshold of sequence similarity constitutes a homolog}? The very existence of sequence similarity between proteins is what makes it possible for biologists to predict unknown interactions and most computational methods rely on it for PPI prediction.

@@

In our experiments section, we thus evaluate our models in two settings - (1) on the original dataset (2) by removing homologous PPI.

>>>>

\abstract{\textbf{Motivation:}
Disease causing pathogens such as viruses, introduce their proteins into the host cells 
where they interact with the host's proteins enabling the virus to replicate inside the host.
These interactions between pathogen and host proteins are key to understanding infectious diseases.
Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. \\
\textbf{Results:}
Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
We obtain between 7 to 39 percentage points improvement in predictive performance over prior state-of-the-art models. 
We show how our model's parameters can be interpreted to reveal both general and specific
interaction-relevant characteristics of the viruses. \\
\textbf{Availability:}
Code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl_mtl.tgz}\\
\textbf{Contact:}
\href{mkshirsa@cs.cmu.edu}{mkshirsa@cs.cmu.edu}
}

@@

\section{Dataset}
\label{sec:datasets}
We use three human-virus PPI datasets from the PHISTO \citep{phisto} database (version from 2014), the characteristics of which are summarized in Table \ref{datasets}. 
The \textit{Influenza A} task includes various strains of flu such as: H1N1 (Puerto Rico, Texas strains etc), H3N2 (England strain etc), H5N1, H7N3.
The \textit{Hepatitis} task has several subtypes (1a, 1b), isolates such as H77, HC-J6 etc and \textit{Ebola} has the Zaire ebolavirus strain Mayinga and strain 1995.
All three are single-strand RNA viruses, with \textit{Hepatitis} being a positive-strand ssRNA whereas \textit{Influenza} and \textit{Ebola} are negative-strand viruses. The density of the known interactions is quite small when considering the entire proteome (i.e all known proteins) of the host and pathogen species (last row in Table \ref{datasets}). \\

\subsection{Addressing the homologs in PHISTO}
\label{sec:phisto}
The interactions data reported in PHISTO database is aggregated from several host-pathogen interactions data sources such as MINT, IntAct, DIP etc. We found that many of the interactions (within a task) involve homologous PPI - virus protein homologs from two strains interacting with either the same human protein or with homologous human proteins. Such PPI usually come from different publications/ studies, but they introduce a lot of redundancy in the dataset. Removing such PPI is challenging as there are several criteria to consider, the most important being: \textit{what threshold of sequence similarity constitutes a homolog}? The very existence of sequence similarity between proteins is what makes it possible for biologists to predict unknown interactions and most computational methods rely on it for PPI prediction.

@@

In our experiments section, we thus evaluate our models in two settings - (1) on the original dataset (2) by creating partitions: the first without any homologous PPI and the other with only the homologous PPI.
